Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-04-05 dictionary.tex
Added to introduction.

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[The function--sheaf dictionary]{The function--sheaf dictionary for smooth commutative group schemes over finite fields}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences, University of Calgary}
\subjclass[2010]{14F05 (primary), 14L15 (secondary)}
\keywords{sheaf--function dictionary, geometrization, quasicharacter sheaves, commutative group schemes}

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes}%showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

Motivated by ...., in this paper we ... The component group of the Neron model of algebraic torus over local field ...   The sheaf--function dictionary for etale group schemes is significantly different . . .\todo{Needs a new front end to really set the tone, heighten the contrast to the connected case and point to applications.}

Let us briefly review how the function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems.
Let $\Fq$ be a finite field and let $G$ be a connected commutative algebraic group over $\Fq$; fix an algebraic closure $\bFq$ of $\Fq$
and let $\bG$ be the base change of $G$ to $\bFq$.
Fix a prime $\ell$ invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character.
Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$.
While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there exists an isomorphism
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The function--sheaf dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

@@

@@

@@

@@

The proof of Proposition~\ref{prop:functorialG} will be given at the end of the next section.

>>>>

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[Lost in Translation: The function--sheaf dictionary]{Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences, University of Calgary}
\subjclass[2010]{14F05 (primary), 14L15 (secondary)}
\keywords{sheaf--function dictionary, geometrization, quasicharacter sheaves, commutative group schemes}

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes, showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

\subsection*{Overview}
Deligne's function--sheaf dictionary translates characters of connected, commutative algebraic groups $G$ over finite fields Fq into certain local systems on G, from which the character of $G(\Fq)$ can be recovered using the action of Frobenius on $G$.
The local systems which appear in this dictionary are easily characterised and often referred to as \emph{character sheaves} on $G$.
The result is a perfect dictionary between characters of $G(\Fq)$ and character sheaves on G, with which the formidable artillery of algebraic geometry can be brought to bear on the study of characters of $G(\Fq)$.
Although this dictionary extends to the case when $G$ is not connected, and even to all smooth commutative group schemes over $\Fq$, the result is not perfect.
In this generality, the correct class of sheaves on G are called \emph{quasicharacter sheaves}; when $G$ is connected and of finite type, quasicharacter sheaves on $G$ specialize to character sheaves on $G$.
While every character of $G(\Fq)$ may be converted, canonically, into a quasicharacter sheaf on G from which the character of $G(\Fq)$ may be recovered using the trace of Frobenius on G, generally there exist non-trivial quasicharacter sheaves on G for which the trace of Frobenius is trivial; these are called \emph{invisible quasicharacter sheaves} on $G$.
Invisible quasicharacter sheaves on G are precisely the quasicharacter sheaves on $G$ that are lost in translation under the function-sheaf dictionary.
Assuming only that $G$ is a smooth commutative group scheme over $\Fq$ with finitely generated component group, this paper gives a simple cohomological description of invisible quasicharacter sheaves on $G$. The most interesting case is that of etale commutative group schemes over $\Fq$.

Our interest in the function--sheaf dictionary for smooth commutative group schemes over $\Fq$, especially etale commutative group schemes, comes from an application to $p$-adic representation theory.
Motivated by ...., in this paper we ... The component group of the Neron model of algebraic torus over local field ...   The sheaf--function dictionary for etale group schemes is significantly different . . .\todo{Needs a new front end to really set the tone, heighten the contrast to the connected case and point to applications.}

\subsection*{Main result}

In order to give the main result of this paper with more precision, let us briefly review how the function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems.
Let $\Fq$ be a finite field and let $G$ be a connected commutative algebraic group over $\Fq$; fix an algebraic closure $\bFq$ of $\Fq$
and let $\bG$ be the base change of $G$ to $\bFq$.
Fix a prime $\ell$ invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character.
Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$.
While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there exists an isomorphism
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The function--sheaf dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

@@

\subsection*{Illustrative examples}

@@

\subsection*{Application}

@@

\subsection*{Structure of the paper}

@@

The proof of Proposition~\ref{prop:functorialG} will be given at the end of Section~\ref{sec:pullback}.
